Global multivariate model learning from hierarchically correlated data

### 2021-02-06 main.tex
Change abstract

\begin{abstract}
The inverse statistical physics infers maximum-entropy models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations. However, in several applications, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree. In turn, empirical averages of observables superpose intrinsic signals related to the equilibrium distribution of the studied system, and spurious historical (or phylogenetic) signals characterizing the specific correlated data-generating process. The naive application of maximum-entropy techniques therefore leads to systematic biases and an effective reduction of the sample size. To advance on the currently open task of extracting intrinsic signals from correlated data, we study a system described by a multivariate Ornstein-Uhlenbeck process defined on a finite tree. Using a Bayesian framework, we can disentangle covariances in the data, which are result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations. Our approach leads to a clear gain in accuracy in the inferred equilibrium distribution, which corresponds to an effective two- to fourfold increase in sample size.
\end{abstract}

@@

\section{Introduction}
\label{sec:int}

With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years. In the most standard setting , the data provide observational samples of the "microscopic" degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc. Within a static modeling approach, frequently based on the maximum-entropy approach, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$. This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system. In some, more rare cases, data correspond to observed time series , but the theoretical and algorithmic development is much less advanced than in the case of independent static data, assuming, e.g., that measurements are taken at the microscopic time scale.

One of the biggest application areas of inverse statistical mechanics is the modeling of biological sequences, and we will use them to motivate our otherwise theoretical-methodological study. Typical datasets are so called homologous protein families, collecting ensembles of proteins of common evolutionary ancestry. Despite their strong divergence in amino-acid sequences, the proteins in such families have highly similar biological functions and three-dimensional fold structures. The different sequences can thereby, in a statistical-physics perspective, be seen as microscopically distinct solutions for the same macroscopic behavior.

In such protein families, which are typically represented by large multiple-sequence alignments, at least two complementary types of biological information are contained:
\begin{itemize}
    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since the common ancestor. Using phylogenetic methods we may reconstruct the evolutionary history of a protein family (and of the species carrying the proteins), and also the sequences of the ancestral proteins. 
    \item {\it Coevolutionary information} considers the correlated evolution of positions in proteins, frequently related to contacts in the folded protein, even for positions distant along the sequence. This information has been massively used in protein-structure prediction, most recently in combination with structurally supervised deep learning, but also to infer mutational landscapes or networks of interacting proteins.
\end{itemize}

These two types of information are contained in two complementary features of the data set: phylogenetic inference is based on the comparative analysis of the rows of the best, i.e., of entire proteins; and coevolutionary inference is based on the comparative analysis of MSA columns, which describe the amino-acid usage of the individual positions across proteins.

For methodological reasons, each type of inference assumes the absence of the other information: phylogenetic inference assumes independent evolution of all positions; coevolutionary inference assumes that proteins form an independently and identically distributed sample of the model to be inferred.

It has turned out to be hard to combine the two types of information in a single approach, even if the existence of, e.g, phylogenetic correlations between sequences influences the results of coevolutionary analysis. Here we propose to advance on the question of phylogeny-aware coevolutionary inference using the simplest nontrivial model setting: correlated multivariate Ornstein-Uhlenbeck (OU) processes.

In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}. 
This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS/EXAMPLES?]. 
A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct models of the selection acting on them. 
A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}.

Models built in this way usually assume that biological characteristics of related organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
This allows simple formulation for model learning. 
However, the phylogenetic relations between evolving entities are in direct contradiction with this simplifying hypothesis, and ignoring them can result in biases when learning models. 
For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.

In order to correct for these biases, it is necessary to disentangle the effects due to selection forces and those due to phylogeny. 
However, this becomes a hard problem when dealing with complex models where traits under study are not independent from each other. 
Accounting for these biases then requires one to understand how interdependent traits under selection evolve along a tree. 
A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
This method is commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}.

This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}. 
In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.

>>>>

\begin{abstract}
Inverse statistical physics aims at inferring models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations. However, in several applications such as biology, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree. In turn, empirical averages of observables superpose intrinsic signals related to the equilibrium distribution of the studied system, and spurious historical (or phylogenetic) signals characterizing the specific correlated data-generating process. The naive application of inverse statistical physics techniques therefore leads to systematic biases and an effective reduction of the sample size. To advance on the currently open task of extracting intrinsic signals from correlated data, we study a system described by a multivariate Ornstein-Uhlenbeck process defined on a finite tree. Using a Bayesian framework, we can disentangle covariances in the data, which are a result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations. Our approach leads to a clear gain in accuracy in the inferred equilibrium distribution, which corresponds to an effective two- to fourfold increase in sample size.
\end{abstract}

@@

\section{Introduction}
\label{sec:int}
% Keep
With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years. 
In the most standard setting, the data provide observational samples of the ``microscopic''degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc... 
Within a static modeling approach, frequently based on the maximum-entropy principle, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$. 
This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system. 
In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than in the case of independent static data. %, assuming, e.g., that measurements are taken at the microscopic time scale.

% Merge with my first par, don't be too specific with protein sequences
One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them. 
A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
We will use this type of application to motivate our otherwise purely methodological study.

% Keep (shorten?)
When modeling homologous genes or organisms, one has to distinguish between two complementary types of information: 
\begin{itemize}
    \item {\it Phylogenetic information:} the distances between gene sequences or traits in organisms carry information about the evolutionary time since their common ancestor. 
    Using phylogenetic methods we may reconstruct the evolutionary history of a set of genes or species, and also the sequences or traits of their ancestors. 
    \item {\it Co-evolutionary information:} positions in a sequence or traits in organisms typically do not evolve independently, but rather in a correlated way. 
    This co-evolution carries important information about the selection forces acting on evolving entities.  
    This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins.  
\end{itemize}

These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different species or genes, while co-evolutionary information is contained in the correlation of different trait values in the same gene or organism.

Models built using biological data usually ignore one of the two types of information. 
For instance, inference of phylogenies typically assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that their sequences are independently distributed. 
This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures. 
However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.

Accounting for both co-evolution and phylogeny in models requires one to understand how interdependent traits under selection evolve along a tree. 
A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
OU processes are commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}. 
This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}. 
In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.
